---
title: "Social Housing in Toronto: Investigating Status"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```


# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....



# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it. 





# Results


The dataset "Centralized Waitlist for Social Housing" provides detailed information on the activities related to the social housing waitlist in Toronto for various quarters starting from Q1 2019. It includes data on new/reactivated applications, the number of households housed, inactive/cancelled applications, and the total active waiting list. Additionally, it breaks down the waiting list by household type (e.g., no dependents, with dependents, seniors).


```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Load required packages
library(ggplot2)

# Extract relevant columns with adjusted column names
year_quarter <- raw_data$Quarter_Year
new_reactivated_applications <- raw_data$New_reactivated
housed <- raw_data$Housed

# Create a data frame
plot_data <- data.frame(Year_Quarter = year_quarter,
                        New_Reactivated_Applications = new_reactivated_applications,
                        Housed = housed)

# Convert Year_Quarter to factor to maintain order
plot_data$Year_Quarter <- factor(plot_data$Year_Quarter, levels = unique(plot_data$Year_Quarter))

# Create a line plot
ggplot(plot_data, aes(x = Year_Quarter)) +
  geom_line(aes(y = New_Reactivated_Applications, color = "New/Reactivated Applications"), size = 1.2) +
  geom_line(aes(y = Housed, color = "Housed"), linetype = "dashed", size = 1.2) +
  scale_color_manual(values = c("blue", "red")) +
  labs(title = "Yearly New/Reactivated Applications vs. Housed",
       x = "Year Quarter",
       y = "Application Status") +
  theme_minimal() +
  theme(legend.title = element_blank())

```
The line graph illustrates the dynamic interplay between demand and supply within Toronto's social housing sector from Q1 2019 to Q4 2023. The data exhibits fluctuating patterns in new or reactivated social housing applications across different quarters, with a notable surge observed in Q1 2020. Despite these fluctuations, the number of households housed each quarter consistently lags behind new entries, underscoring a substantial gap between demand and supply dynamics. This persistent shortfall indicates the ongoing challenge of meeting the housing needs of vulnerable populations within the community.

Examining waiting list trends further illuminates the complexities involved in managing social housing demand. The total active waiting list experiences fluctuations, notably declining in Q2 2019 due to a significant number of inactive or cancelled applications. This fluctuation suggests challenges in maintaining accurate waiting lists or changes in applicants' circumstances, highlighting the multifaceted nature of addressing social housing needs. Such insights underscore the importance of robust data management and adaptive policies to effectively respond to evolving housing demands.

Moreover, an analysis of the waiting list composition reveals critical insights into the demographics of those in need of social housing. A significant portion of seniors and households without dependents feature prominently in the waiting list, indicating a pressing need for housing support among these vulnerable groups. This demographic composition underscores broader social and economic challenges, including affordability issues for seniors and single individuals, necessitating targeted interventions and policy measures to address housing inequities and ensure inclusive access to adequate housing for all segments of the population.


```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```




# Discussion

## First discussion point {#sec-first-point}
The observed dynamics in social housing demand and supply, as evidenced by the fluctuating patterns in application rates and the persistent gap between new entries and households housed, underscore the complex interplay of socio-economic factors influencing housing access and affordability. Research by Desmond and Kimbro (2015) highlights the intricate relationship between housing instability and poverty, wherein marginalized communities often face challenges in accessing stable and affordable housing options. Moreover, studies by Fitzpatrick et al. (2019) emphasize the role of structural inequalities and systemic barriers in perpetuating housing disparities, particularly among vulnerable populations such as seniors and single individuals. The phenomenon of fluctuating waiting list trends and demographic compositions may be attributed to various factors, including inadequate affordable housing stock, rising housing costs, and limited access to supportive services and resources. Additionally, the complexities of managing waiting lists and addressing changing applicant circumstances further contribute to the observed variations in social housing demand. Effective policy interventions aimed at addressing housing inequities and enhancing housing affordability are crucial in mitigating these challenges and ensuring equitable access to housing for all members of society (Desmond & Kimbro, 2015; Fitzpatrick et al., 2019).

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows... 

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```



\newpage


# References


